---
title: "How to inspect your data in an R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

Given a data set, it is very useful to be able to carry out exploratory data analysis on it. This is where we want to see what we have got, what it looks like, whether any data values are missing, and so on.

Let's look first at the `airquality` data set that is built into R

When we load this into R, we get a data frame called airquality

```{r}
data(airquality)
class(airquality)
```

In the case of a data set that comes built-in to R, like airquality, we can find out what it contains by using the usual help facility within R, whereby we type a quation mark in the console, followed by the name of the thing we want help about:

```{r}
?airquality
```

The help information then appears in the Help window in the bottom right-hand pane of RStudio.

In general, where our data does not come from within R, we cannot do this. We can, however, so several things that give us an idea of our data. 

### str()

We can use the `str()`to find out how many rows and columns there are, what kinds of data the columns contain, and what each column is called:

```{r}
str(airquality)
```

In a data frame, you can think of the data as being organised loke in a spreadsheet, with columns of variables, and each row being an observation. So, in the airquality data set, there are 154 observations, where each observation was of 6 variables.

### summary()
To get summary statistics on each variable, we can use summary(). For each variable, this tells us the minimum and maximum values, the mean, the median and the first and third quartile values.

```{r}
summary(airquality)
```



### Missing Values - NAs

Note that for the `Ozone` and `Solar.R` variables a number of 'NA's are listed. These are missing values, which are a very common feature of real data sets. In a typical data analysis, you need to identify whether these are present, and if so, what to do about them. In many cases, R functions contain an additional optional argument `na.rm=TRUE` that allows NAs to be ignored:

Try calculating the mean of the Ozone column:

```{r}
mean(airquality$Ozone)
```
It does not work because of the NA values within the column. However, if we ignore these using the `na.rm=TRUE` argument, all is well:

```{r}
mean(airquality$Ozone, na.rm=TRUE)
```

### head() and tail()
Sometimes it is helpful to actually see a few rows of data. The `head()` and 'tail()` commands reveal the first and last few lines of data, respectively.

```{r}
head(airquality)
```

```{r}
tail(airquality)
```

The default is to reveal 6 rows. If you want see a different number, then you can specify the number of rows to be shown, for example: `head(airquality,10)`.

Let's convert the month  column to a *factor*. That is R's term for a categorical variable. While we are at it, let us give the *levels* of this factor, which are the different values it takes, descriptive labels. These we will take from the built-in R constant `months.abb`, which is a vector of abbreviated names for each month of the year. Lastly we will tell R that these levels are 'ordered', in the sense that June comes after May, etc (whereas, in another context, apples do not come fater lemons, say)

```{r message=FALSE}
airquality$Month <- factor(airquality$Month, levels=5:9, labels=month.abb[5:9], ordered=TRUE)
```

Let's also convert the days column to be an ordered factor, this time with levels 1:31
```{r message=FALSE}
airquality$Day <- factor(airquality$Day, levels=1:31, ordered=TRUE)
```


### Box plot and strip plot
We can plot the ozone level, split by month, using either a `boxplot`

```{r}
boxplot(Ozone ~ Month,data=airquality,ylab="Ozone conc. (ppb)")
```

or, here where there are not many data points, a `stripplot`, from the `lattice` package.
```{r,message=FALSE}
library(lattice)
stripplot(Ozone~ Month,data=airquality,ylab="Ozone conc. (ppb)")
```

### Histogram
We can do a histogram, for example of the wind speed values.

```{r}
histogram(airquality$Wind)
```

If we don't like the default binwidth, we can change it. See ?hist for the various ways this can be done. For exmple here, it might be better to have 25 bins of width 1 mph. We might also prefer another colour.

```{r}
histogram(airquality$Wind,breaks=c(1:25),xlab="Wind speed (mph)",col="grey")
```







