---
title: "How to inspect your data in an R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

Given a data set, it is very useful to be able to carry out exploratory data analysis on it. This is where we want to see what we have got, what it looks like, whether any data values are missing, and so on.

Let's look first at the `airquality` data set that is built into R

When we load this into R, we get a data frame called airquality

```{r}
data(airquality)
class(airquality)
```

In the case of a data set that comes built-in to R, like airquality, we can find out what it contains by using the usual help facility within R, whereby we type a quation mark in the console, followed by the name of the thing we want help about:

```{r}
?airquality
```

The help information then appears in the Help window in the bottom right-hand pane of RStudio.

In general, where our data does not come from within R, we cannot do this. We can, however, so several things that give us an idea of our data. 

### str()

We can use the `str()`to find out how many rows and columns there are, what kinds of data the columns contain, and what each column is called:

```{r}
str(airquality)
```

In a data frame, you can think of the data as being organised loke in a spreadsheet, with columns of variables, and each row being an observation. So, in the airquality data set, there are 154 observations, where each observation was of 6 variables.

### summary()
To get summary statistics on each variable, we can use summary(). For each variable, this tells us the minimum and maximum values, the mean, the median and the first and third quartile values.

```{r}
summary(airquality)
```



### Missing Values - NAs

Note that for the `Ozone` and `Solar.R` variables a number of 'NA's are listed. These are missing values, which are a very common feature of real data sets. In a typical data analysis, you need to identify whether these are present, and if so, what to do about them. In many cases, R functions contain an additional optional argument `na.rm=TRUE` that allows NAs to be ignored:

Try calculating the mean of the Ozone column:

```{r}
mean(airquality$Ozone)
```
It does not work because of the NA values within the column. However, if we ignore these using the `na.rm=TRUE` argument, all is well:

```{r}
mean(airquality$Ozone, na.rm=TRUE)
```

### head() and tail()
Sometimes it is helpful to actually see a few rows of data. The `head()` and 'tail()` commands reveal the first and last few lines of data, respectively.

```{r}
head(airquality)
```

```{r}
tail(airquality)
```

The default is to reveal 6 rows. If you want see a different number, then you can specify the number of rows to be shown, for example: `head(airquality,10)`.

### Factors - categorical variables

Let's convert the month  column to a *factor*. That is R's term for a categorical variable. While we are at it, let us give the *levels* of this factor, which are the different values it takes, descriptive labels. These we will take from the built-in R constant `months.abb`, which is a vector of abbreviated names for each month of the year. Lastly we will tell R that these levels are 'ordered', in the sense that June comes after May, etc (whereas, in another context, apples do not come fater lemons, say)

```{r message=FALSE}
airquality$Month <- factor(airquality$Month, levels=5:9, labels=month.abb[5:9], ordered=TRUE)
```

For the days column, it might be useful to know what day of the week each observation takes place on, since that might help us establish a link between  ozone concentration and ctraffic levels, for example.

To do this, we can create a new column called `dow`, which first has the numbers 1:7 repeating, with '1' indication a Monday. Since the first observation tok place on 1st May 1973, which was a Tuesday, we can achieve our aim like this:

```{r}
airquality$dow<-rep_len(c(2,3,4,5,6,7,1),153)
```

This replicates the sequence 2,3,4,5,6,7,1 until the total length is 153 rows. See `?rep` for further help on this.

Now we can convert this to a labelled, ordered factor, just like we did for the Month column:

```{r message=FALSE}
airquality$dow <- factor(airquality$dow, levels=1:7, labels=c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"), ordered=TRUE)
```

Unlike with the months, there is no built in vector of abbreviations of the days of the week, so we have had to write our own.

### Box plot and strip plot
We can plot the ozone level, split by month or day of the week, using either a `boxplot`

```{r}
boxplot(Ozone ~ Month,data=airquality,ylab="Ozone conc. (ppb)")
boxplot(Ozone ~ dow,data=airquality,ylab="Ozone conc. (ppb)")
```

or, here where there are not many data points, we could also usefully use a `stripplot`, from the `lattice` package.
```{r,message=FALSE}
library(lattice)
stripplot(Ozone~ Month,data=airquality,ylab="Ozone conc. (ppb)")
```

### Histogram
We can do a histogram, for example of the wind speed values.

```{r}
histogram(airquality$Wind)
```

If we don't like the default binwidth, we can change it. See ?hist for the various ways this can be done. For example here, it might be better to have 25 bins of width 1 mph. We might also prefer another colour.

```{r}
histogram(airquality$Wind,breaks=c(1:25),xlab="Wind speed (mph)",col="grey")
```







