---
title: "How to make a table in an R Notebook"
output:
  html_document:
    df_print: paged
---

Suppose we were investigating the `iris` data set that comes with R:

Let us load it into our R session

```{r}
data(iris)
```

and see what we have got:

```{r}
str(iris)
```

To show this as a table, you type:

```
Variable|type|Unit
--------|----|----
Sepal length | numeric | mm
Sepal width | numeric | mm
Petal length | numeric | mm
Petal width | numeric | mm
Species | factor |
```

which, once you knit the file, renders as 

Variable|type|Unit
--------|----|----
Sepal length | numeric | mm
Sepal width | numeric | mm
Petal length | numeric | mm
Petal width | numeric | mm
Species | factor |

For basic tables, that is all there is to it.

### Fancy tables

You can produce really nice tables of data sets in markdown using the package [`kableExtra`](http://haozhu233.github.io/kableExtra/awesome_table_in_html.html)

if you haven't done so already (and if you are reading this, you probably haven't), install the `kableExtra` package and the `knitr` package:

```{r,eval=FALSE}
install.packages("kableExtra")
install.packages("knitr")
```

We load these into our session:

```{r}
library(knitr)
library(kableExtra)
```

Lets try it out on the R data set `mtcars` (it's all about cars). We will take the first 5 rows and the first 6 columns of that and put them into a data frame called dt

```{r}
dt<-mtcars[1:5,1:6]
```

The default table using this package is given like this:

```{r}
kable(dt)
```

The package documentation goes on to show how the style of this table can be tweaked in many ways, if you want to do that.







